pipeComp, a general framework for the evaluation of computational pipelines, reveals performant single cell RNA-seq preprocessing tools

### 2020-02-27 bmc_article.tex

\begin{keyword}
\kwd{single-cell RNAseq}
\kwd{pipeline}
\kwd{clustering}
\kwd{filtering}
\kwd{benchmark}
\end{keyword}

@@

A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq analysis and can guide the choice of methodology (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}). However these recommendations need constant updating and often leave open many details of an analysis. Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow. Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}). A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution\cite{viethSystematic2019,CobosDeconvolution2020}. It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figure 17). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. We therefore favour monitoring several complementary metrics across multiple steps of the process.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth,keepaspectratio]{main_figures/pipeline_explanation.png}
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept of number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function. \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:figure2}
\end{figure}

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \cite{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 2). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \cite{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types). These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs). We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).

We tested \textit{DoubletFinder} \cite{mcginnisDoubletfinder2019} and \textit{scran}'s \texttt{doubletCells} \cite{LunScran2016}, both of which use similarity to artificial doublets, and \textit{scds} \cite{BaisScds2019}, which relies on a combination of co-expression and binary classification. \textit{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \textit{scran} and \textit{scds} return scores that must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets. In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently. We therefore developed a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see Methods).

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/doublets_files/figure-html/figure1-1.png}}
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (DoubletFinder failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \cite{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties. This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Supplementary Figure 1 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 3), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 4). We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 5-7). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.

We next investigated the impact of filtering according to various criteria (see Methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 8). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out. This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).

\begin{figure}
    \centering
    \includegraphics[scale=0.60, keepaspectratio]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{\textbf{A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
    \label{fig:figure4}
\end{figure}

@@

Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}. We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 9). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. An exception was \textit{scVI} (Supplementary Figure TBD) that performed poorly on Zhengmix and simMix1 datasets in view of the subpopulation silhouette and to a relative poor ARI. Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave poorer performance on all metrics. \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{{main_figures/normalization_files/figure-html/figure3-1.png}}
    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' regresses out the number of features and proportion of mitochondrial reads during scaling; `Seurat.feat\_regress` regresses out the number of features only. Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by the ARI at the true number of cluster (left) and by the average mutual information (MI) of the cluster assignment with true subpopulations.}
    \label{fig:figure5}
\end{figure}

Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters. Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 10).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 11A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 11B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 12-13). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 12-14). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 12-13). Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good) but did not perform so well to select genes with a high deviance explained.

\begin{figure}
    \centering
    \includegraphics[scale=0.6]{{main_figures/selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. The asterix denotes the default Seurat method. \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods. Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right). The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
    \label{fig:figure6}
\end{figure}

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 14).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (e.g., Supplementary Figure 15). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics (Figure \ref{fig:figure7}). 
Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 10 and 15). This tended to affect the accuracy of the clustering (Supplementary Figure 16), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

The last step evaluated in our pipeline was clustering. Given previous works on the topic \cite{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNAseq, we restricted our evaluation to Seurat's method and two \textit{scran} SNN-based clustering approaches, based on random walks (\textit{walktrap} method) or on the optimization of the modularity score (\textit{fast\_greedy}). Again, the tested methods were combined with \textit{Seurat}'s standard normalization and \textit{sctransform}, otherwise using the parameters found optimal in the previous steps.

>>>>

\begin{keyword}
\kwd{single-cell RNAseq}
\kwd{pipeline}
\kwd{clustering}
\kwd{normalization}
\kwd{dimensionality}
\kwd{filtering}
\kwd{benchmark}
\end{keyword}

@@

A number of good comparison and benchmark studies have already been performed on various steps related to scRNAseq processing and analysis and can guide the choice of methodology (\citealp{CobosDeconvolution2020, colePerformance2018, DalMolinDE2017, duoClustering2018, freytagComparison2018, GaoProcessing2020, HeiserDimred2019, HouImput2020, JaakkolaDE2017, KrzakClust2019, SonesonDE2018, SunDimRed2019, tianMixology2018, TranBatch2020, TsuyuzakiPCA2020, viethSystematic2019, WangDE2019, YipHVGs2018, ZhangImput2018}). However these recommendations need constant updating and often leave open many details of an analysis. Another missing aspect of current benchmarking studies is their limitation to capture all aspects of scRNAseq processing workflow. Although previous benchmarks already brought valuable recommendations for data processing, some only focused on one aspect of data processing (e.g., \citealp{SunDimRed2019}), did not evaluate how the tool selection affects downstream analysis (e.g., \citealp{TsuyuzakiPCA2020}) or did not tackle all aspects of data processing, such as doublet identification or cell filtering (e.g., \citealp{viethSystematic2019}). A thorough evaluation of the tools covering all major processing steps is however urgently needed as previous benchmarking studies highlighted that a combination of tools can have a drastic impact on downstream analysis, such as differential expression analysis and cell-type deconvolution\cite{viethSystematic2019,CobosDeconvolution2020}. It is then critical to evaluate not only the single effect of a preprocessing method but also its positive or negative interaction with all parts of a workflow.

@@

The \textit{pipeComp} package defines a pipeline as, minimally, a list of functions executed consecutively on the output of the previous one (Figure \ref{fig:figure2}A). In addition, optional benchmark functions can be set for each step to provide standardized, multi-layered evaluation metrics. Given such a \texttt{PipelineDefinition} object, a set of alternative parameters (which might include different subroutines) and benchmark datasets, the \texttt{runPipeline} function then proceeds through all combinations of arguments, avoiding recomputing the same step twice and compiling evaluations (including running time) on the fly. Variations in a given parameter can be evaluated using all metrics from this point downward in the pipeline. This is especially important because end-point metrics, such as the adjusted Rand index (ARI) \cite{HubertARI1985} for clustering, are not perfect. For example, although the meaning of an ARI score is independent of the number of true subpopulations \cite{steinleyProperties2004}, the number of clusters called is by far the most important determinant of the score: the farther it is from the actual number of subpopulations, the worse the ARI (Supplementary Figures 1-2). In this context, one strategy has been to cluster across various resolutions and only consider the results that have the right number of clusters \cite{duoClustering2018}. While this has the virtue of making the results comparable, in practice the number of subpopulations is typically unknown and tools that operate well in this optimal context might not necessarily be best overall. Clustering results are also very sensitive and might not always capture improvements in earlier steps of the pipeline. We therefore favour monitoring several complementary metrics across multiple steps of the process.

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.7, keepaspectratio]{main_figures/pipeline_explanation.png}
    \caption{\textbf{Overview of the \textit{pipeComp} framework and its application to a scRNAseq clustering pipeline. A:} The package is built around a \texttt{PipelineDefinition} S4 class which defines a set of functions to be executed consecutively, as well as optional evaluation functions for each step. Each step can accept a number of parameters whose alternative values are provided as a list. All (or subsets of) combinations of parameters can then be simultaneously ran and evaluated using the \texttt{runPipeline} function. \textbf{B:} Scheme representing the application of \textit{pipeComp} to evaluate a range of methods that are commonly used in scRNAseq studies and some of the metrics monitored at various steps.}
    \label{fig:figure2}
\end{figure}

@@

Doublets, defined as two cells sequenced under the same cellular barcode (e.g., being captured in the same droplet), are fairly frequent in scRNAseq datasets, with estimates ranging from 1 to 10\% depending on the platform and cell concentration used \cite{bloomEstimating2018,kangMultiplexedDemuxlet2018}. While doublets of the same cell type are relatively innocuous for most downstream applications due to their conservation of the relative expression between genes, doublets formed from different cell types or states are likely to be misclassified and could potentially distort downstream analysis. In some cases, doublets can be identified through their unusually high number of reads and detected features, but this is not always the case (Supplementary Figure 3). A number of methods were developed to identify doublets, in most cases by comparing each cell to artificially-created doublets \cite{mcginnisDoubletfinder2019, LunScran2016, BaisScds2019}. We first evaluated the capacity of these methods to detect doublets using the two 10x datasets with cells of different genetic identity \cite{tianMixology2018}, using SNP genotypes as the ground truth. For the sole purpose of this section, we included an additional dataset with SNP information but lacking true cell labels \cite{kangMultiplexedDemuxlet2018}. Of note, SNP-based analyses also call doublets created by cells of the same cell type (but from different individuals) and which are generally described as homotypic (as opposed to neotypic or heterotypic doublets, i.e. doublets from different cell types). These homotypic doublets might not be identifiable from the mere gene counts, and their identification is not generally the primary aim of doublet callers since they are often considered innocuous and, when across individuals, can be identified through other means (e.g., SNPs). We therefore do not expect a perfect accuracy in datasets involving cells of the same type across individuals (as in the demuxlet dataset).

We tested \textit{DoubletFinder} \cite{mcginnisDoubletfinder2019} and \textit{scran}'s \texttt{doubletCells} \cite{LunScran2016}, both of which use similarity to artificial doublets, and \textit{scds} \cite{BaisScds2019}, which relies on a combination of co-expression and binary classification. \textit{DoubletFinder} integrates a thresholding based on the proportion of expected doublets, while \textit{scran} and \textit{scds} return scores that must be manually thresholded. In these cases, we ensured that the right number of cells would be called doublets. In addition to these methods, we reasoned that an approach such as \textit{DoubletFinder} could be simplified by being applied directly on counts and by using a pre-clustering to create neotypic/heterotypic doublets more efficiently. We therefore also evaluated a simple and fast Bioconductor package implementing this method for doublet detection, \textit{scDblFinder}, with the added advantage of accounting for uncertainty in the expected doublet rate and using meta-cells from the clusters to even include triplets (see Methods).

@@

\begin{figure}
    \centering
    \includegraphics[scale=0.65, keepaspectratio]{{main_figures/doublets_files/figure-html/figure1-1.png}}
    \caption{\textbf{Identification of doublet cells. A:} Receiver operating characteristic (ROC) curves of the tested doublet detection methods for three datasets with SNP-identified doublets. Dots indicate threshold determined by the true number of doublets. \textbf{B:} Running time of the different methods (\textit{DoubletFinder} failed on one of the datasets). \textbf{C:} Rate of misclassification of the cells identified by \textit{scDblFinder} as doublets (DBL) or singlets (SNG), across a large range of clustering analysis. Even in datasets which should not have neotypic doublets, the cells identified as such tend to be misclassified.    \label{fig:figure3}}
\end{figure}

@@

Beyond doublets, a dataset might include low-quality cells whose elimination would reduce noise. This has for instance been demonstrated for droplets containing a high content of mitochondrial reads, often as a result of cell degradation and resulting loss of cytoplasmic mRNAs \cite{IlicicLowQual2016}. A common practice is to exclude cells that differ considerably from most other cells on the basis of such properties. This can for instance be performed through the \texttt{isOutlier} function from \textit{scater} that measures, for a given control property, the number of median absolute deviations (MADs) of each cell from the median of all cells. Supplementary Figure 4 shows the distributions of some of the typical cell properties commonly used. Of note, these properties tend to be correlated, with some exceptions. For example, while a high proportion of mitochondrial reads is often correlated with a high proportion of the counts in the top features, there can also be other reasons for an over-representation of highly-expressed features (Supplementary Figure 5), such as an over-amplification in non-UMI datasets. In our experience, 10X datasets also exhibit a very strong correlation between the total counts and the total features even across very different cell types (Supplementary Figure 6). We therefore also measure the ratio between the two and treat cells strongly departing from this trend with suspicion.

Reasoning that the cells we wish to exclude are the cells that would be misclassified, we measured the rate of misclassification of each cell in each dataset across a variety of clustering pipelines, correcting for the median misclassification rate of the subpopulation and then evaluated what properties could be predictive of misclassification (Supplementary Figures 7-9). We could not identify any property or simple combination thereof that would be consistently predictive of misclassification; the only feature that consistently stood out across multiple datasets (the Zheng datasets) was that cells with very high read counts have a higher chance of being misclassified.

We next investigated the impact of filtering according to various criteria (see Methods). An important risk of excluding cells on the basis of their distance from the whole distribution of cells on some properties (e.g., library size) is that these properties tend to have different distributions across subpopulations. As a result, thresholds in terms of number of MADs from the whole distribution can lead to strong biases against certain subpopulations (Figure \ref{fig:figure4}A). We therefore examined the tradeoff between the increased accuracy of filtering and the maximum proportion of cells excluded per subpopulation (Figure \ref{fig:figure4}B and Supplementary Figure 10). Since filtering changes the relative abundance of the different subpopulations, global clustering accuracy metrics such as ARI are not appropriate here. We therefore calculated the per-subpopulation precision and recall using the Hungarian algorithm \cite{PapadimitriouHu1998} and monitored their mean F1 score. A first observation was that, although the stringent filtering tended to be associated with an increase in accuracy, it could also become deleterious and most of the benefits could be achieved without very stringent filtering and minimizing subpopulation bias (Figure \ref{fig:figure4}B). Applying the same filtering criteria on individual clusters of cells (identified through scran's \texttt{quickCluster} method) resulted in nearly no cell being filtered out. This suggests that filtering on the global population tends to discard cells of subpopulations with more extreme properties (e.g. high library size), rather than low-quality cells. Finally, by changing the distributions, the doublet removal step in conjunction with filtering sometimes resulted in a net decrease in the proportion of excluded cells while retaining or improving accuracy. We therefore recommend the use of doublet removal followed by relatively mild filtering, such as is implemented in our `default' filtering (see Methods).

\begin{figure}
    \centering
    \includegraphics[scale=0.60, keepaspectratio]{{main_figures/filtering_files/figure-html/figure2-1.png}}
    \caption{\textbf{Effect of filtering on cell subpopulation structure and clustering. A:} Filtering on the basis of distance to the whole distribution can lead to strong bias against certain subpopulations. The dashed line indicates a threshold of 2.5 median absolute deviations (MADs) from the median of the overall population. \textbf{B:} Relationship between the maximum subpopulation exclusion rate and the average clustering accuracy per subpopulation across various filtering strategies. Of note, doublet removal appears to be desirable even when, due to the design, there are no heterotypic doublets in the data. The PCA methods refer to multivariate outlier detected as implemented in \textit{scater} (see Methods for details).}
    \label{fig:figure4}
\end{figure}

@@

Mitochondrial reads have been associated with cell degradation and there is evidence that ribosomal genes can influence the clustering output, hiding other biological structure in the analysis \cite{freytagComparison2018}. We therefore investigated whether excluding one category of features or the other, or using only protein-coding genes, had an impact on the ability to distinguish subpopulations (Supplementary Figure 11). Removal of ribosomal genes robustly reduced the quality of the clustering, suggesting that they represent real biological differences between subpopulations. Removing mitochondrial genes and restricting to protein-coding genes had a very mild impact.

@@

We further evaluated normalization methods by investigating their impact on the separability of the subpopulations (Figure \ref{fig:figure5}B-C). Since clustering accuracy metrics such as the ARI are very strongly influenced by the number of clusters, we complemented it with silhouette width \cite{RousseeuwSil1987} and mutual information (MI). We found most methods (including no normalization at all) to perform fairly well in most of the subpopulations. An exception was \textit{scVI} (Supplementary Figure 12) which performed poorly on most datasets in terms of both the subpopulation silhouette and clustering accuracy. Scaling tended to reduce the average silhouette width of some subpopulations and to increase that of some less distinguishable ones and was generally, but not always, beneficial on the accuracy of the final clustering. Regressing out covariates systematically gave poorer performance on all metrics. \textit{sctransform} systematically outperformed other methods and, even though it was developed to be applied to data with unique molecular identifiers (UMI), it also performed fairly well with the Smart-seq protocol (Koh and Kumar datasets).

\begin{figure}
    \centering
    \includegraphics[scale=0.55, keepaspectratio]{{main_figures/normalization_files/figure-html/figure3-1.png}}
    \caption{\textbf{Evaluation of normalization procedures.} `Seurat.feat\_mt\_regress' regresses out the number of features and proportion of mitochondrial reads during scaling; `Seurat.feat\_regress` regresses out the number of features only. Residual correlation of the first principal component with library size (left) and detection rate (right), after accounting for biological differences between subpopulations. \textbf{B:} Average silhouette width per true subpopulation, where higher silhouette width means a higher separability. \textbf{C:} Clustering accuracy (\textit{Seurat} clustering), measured by the ARI at the true number of cluster (left) and by the average MI of the cluster assignment with true subpopulations.}
    \label{fig:figure5}
\end{figure}

Finally, we monitored whether, under the same downstream clustering analysis, different normalization methods tended to lead to an over- or under-estimation of the number of clusters. Although some methods had a tendency to lead to a higher (e.g., \textit{sctransform}) or lower (e.g., stable genes) number of clusters, the effect was very mild and not entirely systematic (Supplementary Figure 13). We also confirmed that, especially in the UMI-based datasets, \textit{sctransform} did in fact successfully stabilize variance across mean counts (Supplementary Figure 14).

@@

Reasoning that a selection method should ideally select genes whose variability is higher between subpopulations than within, we first assessed to what extent each method selected genes with a high proportion of variance or deviance explained by (real) subpopulation. As the proportion of variability in a gene attributable to subpopulations can be measured in various ways, we first compared three approaches: ANOVA on log-normalized count, ANOVA on \textit{sctransform} normalization, and deviance explained. The ANOVAs performed on a standard \textit{Seurat} normalization and on \textit{sctransform} data were highly correlated (Supplementary Figure 15A). These estimates were also in good agreement with the deviance explained, although lowly-expressed genes could have a high deviance explained without having much of their variance explained by subpopulation (Supplementary Figure 15B-D). We therefore compared the proportion of the cumulative variance/deviance explained by the top X genes that could be retrieved by each gene ranking method (Supplementary Figures 16-17). We first focused on the first 1000 genes to highlight the differences between methods, although a higher number of selected genes decreased the differences between methods (Supplementary Figures 16-18). The standardized measures of variability were systematically worse than their non-standardized counterparts in selecting genes with a high proportion of variance explained by subpopulation. Regarding the percentage of deviance explained however, the standardized measures were often superior (Figure \ref{fig:figure6}A and Supplementary Figures 16-17). Deviance proved the method of choice to prioritize genes with a high variance explained by subpopulations (with mere expression level proving surprisingly good) but did not perform so well to select genes with a high deviance explained.

\begin{figure}
    \centering
    \includegraphics[scale=0.65]{{main_figures/selection_files/figure-html/figure4-1.png}}
    \caption{\textbf{Evaluation of feature selection methods. A:} Ability of different feature ranking methods to capture genes with a high proportion of variance (left) or deviance (right) explained by real subpopulations. The asterisk denotes the default Seurat method. \textbf{B:} Accuracy of clusterings (at the true number of clusters) when selecting 1000 genes using the given methods. Based on standard \textit{Seurat} normalization (left) or \textit{sctransform} (right). The  \textit{vst.varExp} and \textit{devianceExplained} methods correspond to the estimates used in \textbf{A} to evaluate the selection methods and were included here only for validation purpose.}
    \label{fig:figure6}
\end{figure}

We next evaluated how the use of different feature selection methods affected the clustering accuracy (Figure \ref{fig:figure6}B). To validate the previous assay on the proportion of variance/deviance explained by real populations, we included genes that maximized these two latter measures. Interestingly, while these selections were on average the top-ranking methods, they were not systematically best for all datasets. The previous observations were reflected in the ARI of the resulting clustering (Figure \ref{fig:figure6}B): non-standardized measures of variability, including mere expression level, tended to outperform more complex metrics. In general, we found deviance and unstandardized estimates of variance to provide the best results across datasets and normalization methods. Increasing the number of features selected also systematically led to an increase in the accuracy of the clustering, typically plateauing after 4000 features (Supplementary Figure 18).

@@

The impact of the choice of dimensionality reduction method was far greater than that of normalization or feature selection (Figure \ref{fig:figure7}). GLM-PCA tended to increase the average silhouette width of already well-defined subpopulations, but \textit{Seurat}'s PCA procedure however proved superior on all metrics. Like GLM-PCA, \textit{scVI}'s Linear Decoder (LD) does not explicitly rely on normalized counts data. Its performance was lower than a pipeline with non-weighted \textit{Seurat} PCA on all measures for datasets with more subpopulations (Supplementary Figure 12). 
Overall, weighting the principal components by their variance (as \textit{Seurat} does) had a positive impact on silhouette widths and ARI scores.

@@

A common step following dimension reduction is the selection of an appropriate number of dimensions to use for downstream analysis. Since Euclidean distance decreases as the number of non-discriminating dimensions increases, there is usually a trade-off between selecting enough dimensions to keep most information and excluding smaller dimensions that may represent technical noise or other unwanted sources of variation. Overall, increasing the number of dimensions robustly led to a decrease in the number of clusters (Supplementary Figures 13 and 19). This tended to affect the accuracy of the clustering (Supplementary Figure 20), although in both cases (number of clusters and ARI) \textit{Seurat}'s resolution parameter had a much stronger impact.

@@

The last step evaluated in our pipeline was clustering. Given previous works on the topic \cite{duoClustering2018,freytagComparison2018} and the success of graph-based clustering methods for scRNAseq, we restricted our evaluation to Seurat's method and variations of \textit{scran}'s graph-based clustering using random walks (\textit{walktrap} method) or the optimization of the modularity score (\textit{fast\_greedy}) on various underlying graphs. Specifically, we compared rank-based graphs (\textit{scran}'s default) with graphs using the components' values directly, SNN vs KNN graphs, and the exact nearest neighbors to the Annoy approximation. Again, the tested methods were combined with \textit{Seurat}'s standard normalization and \textit{sctransform}, otherwise using the parameters found optimal in the previous steps.

Since ARI is dominated by differences in the number of clusters (Supplementary Figures 1-2) and no single metric is perfect, we diversified them (Figure \ref{fig:figure9}). MI has the virtue of not decreasing when a true subpopulation is split into two clusters, which is arguably less problematic (and might well reflect unknown biological subgroups), but as a consequence it can be biased towards methods producing higher resolution clustering (Supplementary Figure 2). Similarly, precision per true subpopulation is considerably more robust to differences in the number of clusters. We also tracked the mean F1 score and the ARI at the true number of clusters.
